### 2025-09-19 YEAR2/MAT292_LectureNotes/main.tex
sept19

\begin{example}[Logistic Equation]
    The logistic equation is a first-order nonlinear separable autonomous ODE that models population growth with a carrying capacity. It is given by:
    $$
    \frac{dx}{dt} = \alpha x \left(1 - x \right)
    $$
    where \( \alpha \) is the growth rate.

@@

>>>>

\begin{example}[Logistic Equation] \label{ex:logistic-equation}
    The logistic equation is a first-order nonlinear separable autonomous ODE that models population growth with a carrying capacity. It is given by:
    $$
    \frac{dx}{dt} = \alpha x \left(1 - x \right)
    $$
    where \( \alpha \) is the growth rate.

@@

\subsection{Population Dynamics}
\begin{example}[Simple Model]
    Consider the ODE:
    $$
    P' = rP
    $$
    where \( P(t) \) is the population at time \( t \) and \( r \) is the growth rate. This model assumes that the population grows exponentially without any constraints. The solution to this ODE is:
    $$
    P(t) = P_0 e^{rt}
    $$

    \textit{Is this a good model?} Not really, because it assumes unlimited resources and no environmental constraints, which is unrealistic in real-world scenarios.
\end{example}

\begin{example}[A More Realistic Model]
    Consider the logistic growth model:
        $$
        P' = rP\left(1 - \frac{P}{K}\right)
        $$
        where \( K > 0 \) is the carrying capacity of the environment. This model accounts for limited resources and environmental constraints. When $P = K$, the population growth rate becomes zero, indicating that the population has reached its maximum sustainable size.
        The solution to this ODE is:
        $$
        P(t) = \frac{K}{1 + \left(\frac{K - P_0}{P_0}\right)e^{-rt}}
        $$

        This model is more realistic as it predicts that the population will grow rapidly when small, but will slow down and stabilize as it approaches the carrying capacity \( K \).

        \textbf{Note} This is the same ODE considered in Example \ref{ex:logistic-equation}.

        % Two figures side-by-side using minipages, scaled to the column width
        \begin{figure}[h!]
            \centering
            \begin{minipage}[b]{0.48\textwidth}
                \centering
                \begin{tikzpicture}
                    \begin{axis}[
                        axis lines = middle,
                        xlabel = \( P \),
                        ylabel = \( P' \),
                        ymin = -10, ymax = 10,
                        xmin = -10, xmax = 110,
                        domain=-1:110,
                        samples=200,
                        width=\linewidth,
                        height=5.5cm,
                        grid=both,
                        minor tick num=1,
                        major grid style={line width=.2pt,draw=gray!50},
                        minor grid style={line width=.1pt,draw=gray!20},
                    ]
                    % Logistic growth curve (example parameters r=0.3, K=100)
                    \addplot[blue, thick] {0.3*x*(1 - x/100)};
                    % Fixed points
                    \node at (axis cs:0,0) [circle, draw, inner sep=1.5pt] {};
                    \node at (axis cs:100,0) [circle,fill,inner sep=1.5pt]{};
                    % Labels
                    \node at (axis cs:50,5) [anchor=south west] {\( P' = rP(1 - P/K) \)};
                    \node at (axis cs:0,1) [anchor=south west] {\( P^* = 0 \)};
                    \node at (axis cs:100,1) [anchor=south west] {\( P^* = K \)};
                    \end{axis}
                \end{tikzpicture}
                \vspace{1mm}

                \small\textbf{(a)} Fixed Point Diagram for the Logistic Growth Model
            \end{minipage}\hfill
            \begin{minipage}[b]{0.48\textwidth}
                \centering
                \begin{tikzpicture}
                    \begin{axis}[
                        axis lines = middle,
                        xlabel = \( t \),
                        ylabel = \( P \),
                        ymin = -10, ymax = 150,
                        xmin = 0, xmax = 50,
                        domain=0:50,
                        samples=200,
                        width=\linewidth,
                        height=5.5cm,
                        grid=both,
                        minor tick num=1,
                        major grid style={line width=.2pt,draw=gray!50},
                        minor grid style={line width=.1pt,draw=gray!20},
                    ]
                    % Logistic growth solution curves for different initial conditions (K=100, r=0.3)
                    \addplot[blue, thick] {100/(1 + (100/10 - 1)*exp(-0.3*x))}; % P0 = 10
                    \addplot[blue, thick] {100/(1 + (100/50 - 1)*exp(-0.3*x))}; % P0 = 50
                    \addplot[blue, thick] {100/(1 + (100/90 - 1)*exp(-0.3*x))}; % P0 = 90

                    % carrying capacity line
                    \draw[dashed] (axis cs:0,100) -- (axis cs:50,100);
                    \node at (axis cs:50.5,100) [anchor=west] {\( P = K \)};

                    \addplot[red, thick] {100/(1 + (100/190 - 1)*exp(-0.3*x))}; % P0 = 10
                    \addplot[red, thick] {100/(1 + (100/150- 1)*exp(-0.3*x))}; % P0 = 50
                    \addplot[red, thick] {100/(1 + (100/110 - 1)*exp(-0.3*x))}; % P0 = 90

                    % Labels for initial conditions
                    \node at (axis cs:25,85) [anchor=south west] {\( P_0 = 10 \)};
                    \node at (axis cs:25,60) [anchor=south west] {\( P_0 = 50 \)};
                    \node at (axis cs:25,40) [anchor=south west] {\( P_0 = 90 \)};
                    \end{axis}
                \end{tikzpicture}
                \vspace{1mm}

                \small\textbf{(b)} Sample solution trajectories (different \(P_0\))
            \end{minipage}
            \caption{Logistic growth: (a) fixed-point diagram; (b) sample solution trajectories for different initial conditions.}
            \label{fig:logistic-growth-side-by-side}
        \end{figure}
\end{example}

\begin{example}[Logistics with Threshold]
    Now we introduce a threshold \( T \) below which the population cannot sustain itself:
        $$
        P' = rP\left(1 - \frac{P}{K}\right)\left(\frac{P}{T} - 1\right)
        $$
        This model introduces an Allee effect, where the population growth rate becomes negative when the population is below the threshold \( T \). The fixed points of this system are:
        $$
        P^* = 0, \quad P^* = T, \quad P^* = K
        $$
        The stability of these fixed points can be analyzed as follows:
        \begin{itemize}
            \item \( P^* = 0 \) is stable (since \( f'(0) < 0 \))
            \item \( P^* = T \) is unstable (since \( f'(T) > 0 \))
            \item \( P^* = K \) is stable (since \( f'(K) < 0 \))
        \end{itemize}

    A phase plot and sample solution trajectories are shown below (the sketch shows guides at \(P=T\) and \(P=K\), and sample trajectories that converge to \(K\) or diverge to \(0\)):

        \begin{figure}[h!]
            \centering
            \begin{minipage}[b]{0.48\textwidth}
                \centering
                \begin{tikzpicture}
                    \begin{axis}[
                        axis lines = middle,
                        xlabel = \( P \),
                        ylabel = \( P' \),
                        ymin = -10, ymax = 10,
                        xmin = -5, xmax = 110,
                        domain=-1:110,
                        samples=200,
                        width=\linewidth,
                        height=5.5cm,
                        grid=both,
                        minor tick num=1,
                        major grid style={line width=.2pt,draw=gray!50},
                        minor grid style={line width=.1pt,draw=gray!20},
                    ]
                    % Logistic growth with threshold curve (example parameters r=0.3, K=100, T=40)
                    \addplot[blue, thick] {0.3*x*(1 - x/100)*(x/40 - 1)};
                    % Axis baseline and vertical guides at P=T and P=K
                    \addplot[black] coordinates {(0,0) (110,0)}; % P-axis
                    \draw[red,dashed]  (axis cs:40,-9) -- (axis cs:40,9) node[black, above right] {\(P=T\)};
                    \draw[teal,dashed] (axis cs:100,-9) -- (axis cs:100,9) node[black, above right] {\(P=K\)};
                    % Fixed points markers on P-axis (at P'=0)
                    \node at (axis cs:0,0) [circle,fill,inner sep=1.5pt,label=below:{\(0\)}] {};
                    \node at (axis cs:40,0) [circle,draw,fill=white,inner sep=1.5pt,label=below:{\(T\)}] {};
                    \node at (axis cs:100,0) [circle,fill,inner sep=1.5pt,label=below:{\(K\)}] {};
                    % Flow direction arrows on P-axis (qualitative)
                    \draw[->,thick] (axis cs:20,0) -- (axis cs:5,0);    % 0<T: P decreases (left)                    \draw[->,thick] (axis cs:45,0) -- (axis cs:70,0.6);   % T<K: P increases (right)
                    \draw[->,thick] (axis cs:75,0) -- (axis cs:95,0);   % >K: P decreases (left)

                    \end{axis}
                \end{tikzpicture}
                \vspace{1mm}    
                \small\textbf{(a)} Phase diagram (fixed points and flow)
            \end{minipage}\hfill
            \begin{minipage}[b]{0.48\textwidth}
                \centering
                \begin{tikzpicture}
                    \begin{axis}[
                        axis lines = middle,
                        xlabel = \( t \),
                        ylabel = \( P \),
                        ymin = -5, ymax = 110,
                        xmin = 0, xmax = 50,    
                        domain=0:50,
                        samples=200,
                        width=\linewidth,
                        height=5.5cm,
                        grid=both,
                        minor tick num=1,   
                        major grid style={line width=.2pt,draw=gray!50},
                        minor grid style={line width=.1pt,draw=gray!20},
                    ]
                    % carrying capacity line
                    \draw[dashed] (axis cs:0,100) -- (axis cs:50,100);
                    \node at (axis cs:50.5,100) [anchor=west] {\( P = K \)};
                    % threshold line (visual)
                    \draw[dashed] (axis cs:0,40) -- (axis cs:50,40);
                    \node at (axis cs:50.5,40) [anchor=west] {\( P = T \)};
                    % Sample solution trajectories (qualitative sketches)
                    % Below-threshold -> decline to 0 (four examples)
                    \addplot[red, thick, opacity=0.9, domain=0:50, samples=200]
                        {40/(1 + (40/35 - 1)*exp(0.25*x))};
                    \addplot[red, thick, opacity=0.9, domain=0:50, samples=200]
                        {40/(1 + (40/25 - 1)*exp(0.25*x))};
                    \addplot[red, thick, opacity=0.9, domain=0:50, samples=200]
                        {40/(1 + (40/15 - 1)*exp(0.25*x))};
                    \addplot[red, thick, opacity=0.9, domain=0:50, samples=200]
                        {40/(1 + (40/5  - 1)*exp(0.25*x))};
                    \node[red] at (axis cs:7,26) [anchor=west] {\(P_0<T\) diverges to $0$};
                    % Above-threshold -> grow to K (four examples)
                    \addplot[blue, thick, opacity=0.9, domain=0:50, samples=200]
                        {100/(1 + (100/45 - 1)*exp(-0.3*x))};
                    \addplot[blue, thick, opacity=0.9, domain=0:50, samples=200]
                        {100/(1 + (100/60 - 1)*exp(-0.3*x))};
                    \addplot[blue, thick, opacity=0.9, domain=0:50, samples=200]
                        {100/(1 + (100/80 - 1)*exp(-0.3*x))};
                    \addplot[blue, thick, opacity=0.9, domain=0:50, samples=200]
                        {100/(1 + (100/95 - 1)*exp(-0.3*x))};
                    \node[blue] at (axis cs:40,82) [anchor=east] {\(P_0>T\) converges to $K$};
                    \end{axis}
                \end{tikzpicture}
                \vspace{1mm}    
                \small\textbf{(b)} Sample time trajectories: convergence to \(K\) or collapse to \(0\)
            \end{minipage}
            \caption{Logistic growth with threshold: (a) phase portrait with horizontal guides at \(P=T\) and \(P=K\); (b) qualitative solution trajectories showing convergence and divergence relative to the threshold.}
            \label{fig:logistic-growth-threshold-side-by-side}
        \end{figure}
\end{example}

\begin{example}[Harvesting Model]
    Consider a population model with harvesting:
    $$
    P' = rP\left(1 - \frac{P}{K}\right) - H
    $$
    where \( H \) is the constant harvesting rate. The fixed points of this system are given by solving:
    $$
    rP\left(1 - \frac{P}{K}\right) - H = 0
    $$
    This leads to a quadratic equation in \( P \):
    $$
    P^2 - KP + \frac{KH}{r} = 0
    $$
    The solutions to this equation are:
    $$
    P^* = \frac{K \pm \sqrt{K^2 - 4\frac{KH}{r}}}{2}
    $$
    The stability of these fixed points can be analyzed as follows:
    \begin{itemize}
        \item If \( H < \frac{rK}{4} \), there are two real fixed points: one stable and one unstable.
        \item If \( H = \frac{rK}{4} \), there is one real fixed point (a saddle-node bifurcation point).
        \item If \( H > \frac{rK}{4} \), there are no real fixed points, indicating that the population will decline to extinction.
    \end{itemize}
    \textbf{Fixed Point Diagram} A fixed point diagram is shown below:
        \begin{figure}[h!]
            \centering
            \begin{tikzpicture}
                \begin{axis}[
                    axis lines = middle,
                    xlabel = \( y \),
                    ylabel = \( f(y) \),
                    ymin = -10, ymax = 10,
                    xmin = -5, xmax = 105,
                    domain=-1:105,
                    samples=200,
                    width=0.6\linewidth,
                    height=6cm,
                    grid=both,
                    minor tick num=1,
                    major grid style={line width=.2pt,draw=gray!50},
                    minor grid style={line width=.1pt,draw=gray!20},
                    legend pos=north east,
                ]
                % Quadratic curve for different H values with legend entries
                \addplot[blue, thick] {0.3*x*(1 - x/100) - 5};
                \addlegendentry{\(H < \tfrac{rK}{4}\)}
                \addplot[red, thick, dashed] {0.3*x*(1 - x/100) - 7.5};
                \addlegendentry{\(H = \tfrac{rK}{4}\)}
                \addplot[green, thick, dotted] {0.3*x*(1 - x/100) - 10};
                \addlegendentry{\(H > \tfrac{rK}{4}\)}
                % Fixed points using plot marks (uses built-in plot labeling via legend)
                \addplot[black, only marks, mark=o, mark size=2.5pt] coordinates {(20,0)};
                \addplot[black, only marks, mark=*, mark size=2.5pt] coordinates {(80,0)};
                % semi-filled marker for intermediate / degenerate fixed point
                % half-filled marker: right half black, left half white
                \node at (axis cs:50,0) [circle, draw, fill=white, inner sep=1.5pt] {};
                \end{axis}
            \end{tikzpicture}
            \caption{Fixed point diagram for the harvesting model with different harvesting rates \(H\).}
        \end{figure}

     \textbf{Bifurcation Diagram} This is a classic example of a saddle-node bifurcation (0 to 2 fixed points). The bifurcation point is at $(k/2, 4h/K)$. The bifurcation diagram is shown below:
        \begin{figure}[h!]
            \centering
            \begin{tikzpicture}
                \begin{axis}[
                    axis lines = middle,
                    xlabel = \( r \),
                    ylabel = \( P^* \),
                    ymin = -10, ymax = 110,
                    xmin = -1, xmax = 30,
                    domain=0:30,
                    samples=200,
                    width=0.6\linewidth,
                    height=6cm,
                    grid=both,
                    minor tick num=1,
                    major grid style={line width=.2pt,draw=gray!50},
                    minor grid style={line width=.1pt,draw=gray!20},
                ]
                % Bifurcation curve (example parameters r=0.3, K=100)
                \addplot[blue, thick, domain=7.5:30] {50 + 10*sqrt(x-7.5)};
                \addplot[red, thick, dashed, domain=7.5:30] {50 - 10*sqrt(x-7.5)};
                \draw[dashed] (axis cs:7.5,-10) -- (axis cs:7.5,110);
                \node at (axis cs:7.5,110) [anchor=north west] {\( r = \frac{4H}{K} \)};
                % Mark bifurcation point
                \node at (axis cs:7.5,50) [circle,fill,inner sep=1.5pt]{};
                \end{axis}
            \end{tikzpicture}
            \caption{Bifurcation diagram for the harvesting model showing the saddle-node bifurcation at \(H = \frac{rK}{4} \).}
        \end{figure}

\begin{theorem}[ID Autonomous ODEs are Monotonic]
    Consider the ODE:
    $$
    y' = f(y)
    $$
    where \( f(y) \) is continuous. Then, the solution \( y(t) \) is monotonic on its interval of existence.

\end{theorem}
\begin{proof}
    WLOG, suppose \( f(y_0) > 0 \). For the sake of contradiction, suppose \( y(t) \) is not monotonic. Then, there exists \( t_1 > t_0 \) such that \( y(t_1) < y(t_0) \). By the Intermediate Value Theorem, there exists \( t_2 \in (t_0, t_1) \) such that \( y(t_2) = y(t_0) \). By the Mean Value Theorem, there exists \( c \in (t_0, t_2) \) such that:
    $$\frac{y(t_2) - y(t_0)}{t_2 - t_0} = y'(c)$$
    But \( y(t_2) = y(t_0) \), so the left-hand side is zero. Thus, \( y'(c) = 0 \). However, since \( f(y) \) is continuous and \( f(y_0) > 0 \), there exists a neighborhood around \( y_0 \) where \( f(y) > 0 \). This contradicts the assumption that \( y'(c) = 0 \). Therefore, \( y(t) \) must be monotonic.
\end{proof}

\begin{definition}[System of ODEs]
    A system of ODEs is a set of coupled first-order ODEs of the form:
    $$
    \begin{cases}
    x_1' = f_1(t, x_1, x_2, \ldots, x_n) \\
    x_2' = f_2(t, x_1, x_2, \ldots, x_n) \\
    \vdots \\
    x_n' = f_n(t, x_1, x_2, \ldots, x_n)
    \end{cases}
    $$
    where \( x_i(t) \) are the state variables and \( f_i(t, x_1, x_2, \ldots, x_n) \) are given functions.
\end{definition}
\subsection{Systems of ODEs}

\begin{definition}[Two-Species Interaction Model]
    Consider a model of two species interacting, such as predator-prey dynamics:
    $$
    \begin{cases}
    x' = x(r_1 - ax -by) \\
    y' = y(r_2 + cx - dy)
    \end{cases}
    $$
    where \( x(t) \) and \( y(t) \) are the populations of the two species, \( r_1, r_2, a, b, c, d \) are constants representing:
    \begin{itemize}
        \item \( r_1, r_2 \): intrinsic growth rates of species \( x \) and \( y \)
        \item \( a, d \): intrinsic carrying capacity coefficients
        \item \( b, c \): interspecific coefficients representing the effect of one species on the other
    \end{itemize}
\end{definition}

\begin{example}[Lotka-Volterra Model]
    Now, consider the classic Lotka-Volterra predator-prey model:
    $$
    \begin{cases}
    x' &= ax - bxy \\
    y' &= -cy + dxy \\
    (x(0), y(0)) &= (x_0, y_0)
    \end{cases}
    $$
    where \( x(t) \) is the prey population, \( y(t) \) is the predator population, and \( a, b, c, d \) are positive constants representing:
    \begin{itemize}
        \item \( a \): growth rate of prey in the absence of predators
        \item \( b \): predator efficiency (how many prey are eaten per predator per unit time)
        \item \( c \): natural death rate of predators in the absence of prey
        \item \( d \): conversion efficiency (how many predators are born per prey eaten)
    \end{itemize}
    Consider the fixed points of this system by setting \( x' = 0 \) and \( y' = 0 \). Then we have:
    $$
    (x^*, y^*) = (0, 0) \quad \text{and} \quad (x^*, y^*) = \left(\frac{c}{d}, \frac{a}{b}\right)
    $$
    The case where \( (x^*, y^*) = (0, 0) \) is trivial (extinction of both species). The non-trivial fixed point \( \left(\frac{c}{d}, \frac{a}{b}\right) \) represents a coexistence equilibrium where the rate of the prey being eaten is the rate of the prey being born.
\end{example}

We can further analyze this by considering Nullclines.

\begin{definition}[Nullclines]
    Nullclines are curves in the phase plane where the derivative of \textbf{one of the variables} is zero. For the Lotka-Volterra model, the nullclines are given by:
    \begin{itemize}
        \item Prey nullcline (\( x' = 0 \)): \( y = \frac{a}{b} \) (horizontal line)
        \item Predator nullcline (\( y' = 0 \)): \( x = \frac{c}{d} \) (vertical line)
    \end{itemize}
    The intersection of these nullclines gives the fixed points of the system.

\begin{example}[Nullclines of the Lotka-Volterra Model]
    The x-nullclines ($x' = 0$) are given by:
    $$x' = ax - bxy = 0 \implies x(a - by) = 0 \implies x = 0 \text{ or } y = \frac{a}{b}$$
    The y-nullclines ($y' = 0$) are given by:
    $$y' = -cy + dxy = 0 \implies y(-c + dx) = 0 \implies y = 0 \text{ or } x = \frac{c}{d}$$

    Also, consider the growth of each species in the nullcline regions:
    \begin{itemize}
        \item For \( x' = 0 \) (prey nullcline):
        \begin{itemize}
            \item If \( x = 0 \), then \( y' < 0 \) (predator population decreases).
            \item If \( y < \frac{a}{b} \), then \( x' > 0 \) (prey population increases).
            \item If \( y > \frac{a}{b} \), then \( x' < 0 \) (prey population decreases).
        \end{itemize}
        \item For \( y' = 0 \) (predator nullcline):
        \begin{itemize}
            \item If \( y = 0 \), then \( x' > 0 \) (prey population increases).
            \item If \( x < \frac{c}{d} \), then \( y' < 0 \) (predator population decreases).
            \item If \( x > \frac{c}{d} \), then \( y' > 0 \) (predator population increases).
        \end{itemize}
    \end{itemize}

    These directions would ultimately leads to a circular flow around the fixed point \( \left(\frac{c}{d}, \frac{a}{b}\right) \).

    The intersection of these nullclines gives the fixed points of the system. A phase portrait showing the nullclines and the fixed point is illustrated below:
        \begin{figure}[h!]
            \centering
            \begin{tikzpicture}
                \begin{axis}[
                    axis lines = middle,
                    view     = {0}{90},
                    xlabel = \( x \) (Prey),
                    ylabel = \( y \) (Predator),
                    ymin = -1, ymax = 6,
                    xmin = -1, xmax = 6,
                    domain=-1:6,
                    samples=200,
                    width=0.6\linewidth,
                    height=6cm,
                    grid=both,
                    minor tick num=1,
                    major grid style={line width=.2pt,draw=gray!50},
                    minor grid style={line width=.1pt,draw=gray!20},
                ]
                % Prey nullcline (horizontal line)
                \addplot[red, thick] {3};
                % trvial nullclines (axes)
                \addplot[blue, thick] {0};
                \node at (axis cs:5,3) [anchor=south west] {\( y = \frac{a}{b} \)};
                % Predator nullcline (vertical line)
                \addplot[red, thick] coordinates {(0,0) (0,6)};
                \addlegendentry{Predator nullcline}
                % trvial nullclines (axes)
                \addplot[blue, thick] coordinates {(2,0) (2,6)};
                \addlegendentry{Prey nullcline}

                \node at (axis cs:2,5) [anchor=north east] {\( x = \frac{c}{d} \)};
                % Fixed point marker
                \node at (axis cs:2,3) [circle,fill,inner sep=1.5pt,label=above right:{\(\left(\frac{c}{d}, \frac{a}{b}\right)\)}] {};
                \end{axis}
            \end{tikzpicture}
            \caption{Nullclines of the Lotka-Volterra model showing the prey nullcline (horizontal line) and predator nullcline (vertical line). The intersection point represents the coexistence equilibrium.}
        \end{figure}

\begin{example}[Simple Pendulum]
    Consider a simple pendulum of length \( L \) and mass \( m \) under the influence of gravity \( g \). The angle \( \theta(t) \) that the pendulum makes with the vertical satisfies the second-order ODE:
    $$
    \theta'' + \frac{g}{L} \sin(\theta) = 0
    $$
    To convert this into a system of first-order ODEs, we let:
    $$\begin{cases}
    x := \theta  \quad x' = \theta' = y \\
    y := \theta' \quad y' = \theta'' = -\frac{g}{L} \sin(x)
    \end{cases}$$

    Consider the fixed points of this system by setting \( x' = 0 \) and \( y' = 0 \). Then we have:
    $$(x^*, y^*) = (n\pi, 0) \quad n \in \mathbb{Z}$$
    The case where \( n \) is even (e.g., \( 0, 2, 4, \ldots \)) corresponds to the pendulum hanging straight down (stable equilibrium), while the case where \( n \) is odd (e.g., \( 1, 3, 5, \ldots \)) corresponds to the pendulum being inverted (unstable equilibrium).

    Consider the nullclines:
    \begin{itemize}
        \item \( x' = 0 \) (horizontal nullcline): \( y = 0 \)
        \item \( y' = 0 \) (vertical nullcline): \( \sin(x) = 0 \implies x = n\pi, n \in \mathbb{Z} \) And we have, for $y$:

    \end{itemize}
\end{example}

### 2025-09-19 YEAR2/PHY293_LectureNotes/main.tex
sept19

\paragraph{The physical pendulum} For a rigid body swinging about a pivot point, the equation of motion is given by:
\begin{equation}
    I\ddot{\theta} = \tau 
\end{equation}
For simple rod of length \( L \) and mass \( m \) pivoted at one end, the moment of inertia about the pivot point is \( I = \frac{1}{3}mL^2 \). The torque due to gravity when the rod is displaced by an angle \( \theta \) from the vertical is \( \tau = -mg\frac{L}{2}\sin(\theta) \). For small angles, we can approximate \( \sin(\theta) \approx \theta \), leading to:
\begin{equation}
    \ddot{\theta} + \frac{3g}{2L}\theta = 0
\end{equation}
The angular frequency of oscillation for the physical pendulum is:
\begin{equation}
    \omega = \sqrt{\frac{3g}{2L}}
\end{equation}

@@

\paragraph{The RLC Circuit} An RLC circuit consists of a resistor \( R \), inductor \( L \), and capacitor \( C \) connected in series. The charge \( q(t) \) on the capacitor satisfies the differential equation:
\begin{equation}
    L\frac{d^2q}{dt^2} + R\frac{dq}{dt} + \frac{1}{C}q = 0
\end{equation}
This is analogous to the resistor behaving as a damped harmonic oscillator, with the angular frequency given by:
\begin{equation}
    \omega = \sqrt{\frac{1}{LC} - \left(\frac{R}{2L}\right)^2}
\end{equation}

\section{The Wave Equation}

>>>>

\paragraph{The physical pendulum} For a rigid body swinging about a pivot point, the equation of motion is given by:
\begin{equation}
    I\ddot{\theta} = \tau 
\end{equation}
For simple rod of length \( L \) and mass \( m \) pivoted at one end, the moment of inertia about the pivot point is \( I = \frac{1}{3}mL^2 \). The torque due to gravity when the rod is displaced by an angle \( \theta \) from the vertical is \( \tau = -mg\frac{L}{2}\sin(\theta) \). For small angles, we can approximate \( \sin(\theta) \approx \theta \), leading to:
\begin{equation}
    \ddot{\theta} + \frac{3g}{2L}\theta = 0
\end{equation}
And the small angle approximates leads to:
\begin{equation}
    \ddot{\theta} + \omega^2 \theta = 0
\end{equation}
where \( \omega = \sqrt{\frac{3g}{2L}} \). The solution to this ODE is:
\begin{equation}
    \theta(t) = \theta_0 \cos(\omega t + \phi)
\end{equation}
The angular frequency of oscillation for the physical pendulum is:
\begin{equation}
    \omega = \sqrt{\frac{3g}{2L}}
\end{equation}

@@

\paragraph{The RLC Circuit} An RLC circuit consists of a resistor \( R \), inductor \( L \), and capacitor \( C \) connected in series. The charge \( q(t) \) on the capacitor satisfies the differential equation:
\begin{equation}
    L\frac{d^2q}{dt^2} + R\frac{dq}{dt} + \frac{1}{C}q = 0
\end{equation}
This is analogous to the resistor behaving as a damped harmonic oscillator, with the angular frequency given by:
\begin{equation}
    \omega = \sqrt{\frac{1}{LC} - \left(\frac{R}{2L}\right)^2}
\end{equation}
The following are equations that summarize the analogies between mechanical and electrical oscillators:
\begin{table}[h!]
    \centering
    \begin{tabular}{|c|c|c|}
        \hline
        \textbf{Mechanical Oscillator} & \textbf{Electrical Oscillator} & \textbf{Analogy} \\
        \hline
        Mass \( m \) & Inductance \( L \) & Inertia \\
        Spring constant \( k \) & Inverse capacitance \( \frac{1}{C} \) & Restoring force \\
        Damping coefficient \( b \) & Resistance \( R \) & Energy dissipation \\
        Displacement \( x \) & Charge \( q \) & Position \\
        Velocity \( \dot{x} \) & Current \( I = \frac{dq}{dt} \) & Rate of change of position \\
        Force \( F \) & Voltage \( V \) & Driving force \\
        \hline
    \end{tabular}
    \caption{Analogies between Mechanical and Electrical Oscillators}
    \label{tab:oscillator_analogies}
\end{table}

\subsection{Coupled Oscillators}
\begin{definition}[Coupled Oscillators]
    Consider two masses \( m_1 \) and \( m_2 \) on two pendulum connected by springs with spring constant \( k \).  The equations of motion for the two masses are given by:

    \begin{equation}
        \begin{cases}
            \ddot{x}_1 &= -\frac{g}{L}x_1 + \frac{k}{m_1} (x_1 - x_2) \quad (a) \\
            \ddot{x}_2 &= -\frac{g}{L}x_2 - \frac{k}{m_2} (x_1 - x_2) \quad (b)
        \end{cases}
    \end{equation}
    where \( x_1 \) and \( x_2 \) are the displacements of masses \( m_1 \) and \( m_2 \) from their equilibrium positions, respectively. Now if we add and subtract the two equations, we get:
    $$
        \begin{cases}
            \ddot{x}_1 + \ddot{x}_2 &= -\frac{g}{L}(x_1 + x_2) \quad (c) \\
            \ddot{x}_1 - \ddot{x}_2 &= -\left(\frac{g}{L} + \frac{k}{m_1} + \frac{k}{m_2}\right)(x_1 - x_2) \quad (d)
        \end{cases}
    $$
    Let:
    $$ 
    q_1 = x_1 + x_2 \quad \text{and} \quad q_2 = x_1 - x_2
    $$
    also, we let:
    $$
        \omega_1 = \sqrt{\frac{g}{L}} \quad \text{and} \quad \omega_2 = \sqrt{\frac{g}{L} + \frac{k}{m_1} + \frac{k}{m_2}}
    $$
    Then, the equations of motion can be rewritten as:
    $$
        \begin{cases}
            \ddot{q}_1 &= -\omega_1^2 q_1 \\
            \ddot{q}_2 &= -\omega_2^2 q_2
        \end{cases}
    $$
    Observe that \( q_1 \) and \( q_2 \) are decoupled, and each behaves like a simple harmonic oscillator with angular frequencies \( \omega_1 \) and \( \omega_2 \), respectively. The general solutions for \( q_1(t) \) and \( q_2(t) \) are:
    \begin{equation}
        \begin{cases}
            q_1(t) &= A_1 \cos(\omega_1 t + \phi_1) \\
            q_2(t) &= A_2 \cos(\omega_2 t + \phi_2)
        \end{cases}
    \end{equation}
    where \( A_1, A_2, \phi_1, \) and \( \phi_2 \) are constants determined by the initial conditions. Finally, we can expressions for \( x_1(t) \) and \( x_2(t) \) in terms of \( q_1(t) \) and \( q_2(t) \):
    \begin{equation}
        \begin{cases}
            x_1(t) &= \frac{q_1(t) + q_2(t)}{2} = \frac{A_1}{2} \cos(\omega_1 t + \phi_1) + \frac{A_2}{2} \cos(\omega_2 t + \phi_2) \\
            x_2(t) &= \frac{q_1(t) - q_2(t)}{2} = \frac{A_1}{2} \cos(\omega_1 t + \phi_1) - \frac{A_2}{2} \cos(\omega_2 t + \phi_2)
        \end{cases}
    \end{equation}
    All parts of the system oscillate with the same normal frequency (eigenvalue) in a normal mode (eigenvector).
\end{definition}

\begin{definition}[Normal Modes]
    A \textbf{normal mode} is a natural pattern of oscillation in which every part of the system moves in sync with a single frequency.

    For the two-mass pendulum system, there are two normal modes:
    \begin{itemize}
        \item \textbf{In-phase mode} (\(\omega_1\)): both masses swing together, reaching maximum displacement on the same side at the same time.  
        $$
        x_1 = x_2
        $$

        \item \textbf{Out-of-phase mode} (\(\omega_2\)): the two masses swing in opposite directions, so when one moves left, the other moves right.  
        $$
        x_1 = -x_2
        $$
    \end{itemize}

    Any general motion of the system can be written as a combination (superposition) of these two normal modes.
\end{definition}
\paragraph{Energy in Coupled Oscillators} The total mechanical energy \( E \) of the coupled oscillator system is given by:
\begin{equation}
    E = \frac{1}{2}m_1\dot{x}_1^2 + \frac{1}{2}m_2\dot{x}_2^2 + \frac{1}{2}k(x_1 - x_2)^2 + \frac{1}{2}m_1\frac{g}{L}x_1^2 + \frac{1}{2}m_2\frac{g}{L}x_2^2
\end{equation}
Substituting the expressions for \( x_1 = \frac{q_1 + q_2}{2} \) and \( x_2 = \frac{q_1 - q_2}{2} \) into the energy equation, two independent energies terms for $q_1$ and $q_2$ emerge:
\begin{equation}
    E = \left[ \frac{1}{4} m \dot{q}_1^2 + \frac{1}{4}\frac{m_1 g}{L} q_1^2 \right] + \left[ \frac{1}{4} m \dot{q}_2^2 + \frac{1}{4}\frac{m_2 g}{L} q_2^2 \right]
\end{equation}
This shows that the total energy is the sum of the energies associated with each normal mode, and that no cross term between the modes exists. Thus energy does not flow from one mode to another.

\begin{example}
    Given two coupled pendulums with $x_0(0) = A$, $x_1(0) = 0$, and both masses at rest initially, we have:
    \begin{align*}
        q_0(t) &= C_0 \cos(\omega_0 t) \\
        q_1(t) &= C_1 \cos(\omega_1 t)
    \end{align*}
    where \( C_0 \) and \( C_1 \) are constants determined by the initial conditions. Using the initial conditions. Using the cosine addition formula, which is:
    \begin{align*}
        \cos(A) + \cos(B) &= 2\cos\left(\frac{A+B}{2}\right)\cos\left(\frac{A-B}{2}\right) \\
        \cos(A) - \cos(B) &= -2\sin\left(\frac{A+B}{2}\right)\sin\left(\frac{A-B}{2}\right)
    \end{align*}
    we can express the motion of each pendulum as:
    \begin{align*}
        x_0(t) &= \frac{q_0 + q_1}{2} = A \cos\left(\frac{\omega_0 + \omega_1}{2} t\right) \cos\left(\frac{\omega_0 - \omega_1}{2} t\right) \\
        x_1(t) &= \frac{q_0 - q_1}{2} = A \sin\left(\frac{\omega_0 + \omega_1}{2} t\right) \sin\left(\frac{\omega_0 - \omega_1}{2} t\right)
    \end{align*}
    Due to the formula for the product of cosines and sines, we see that energy oscillates between the two pendulums with a beat frequency of \( \frac{\omega_1 - \omega_0}{2} \). The time for all the energy to transfer from one pendulum to the other is given by:
    \begin{align*}
        T &= \frac{2\pi}{\frac{\omega_1 - \omega_0}{2}} = \frac{4\pi}{\omega_1 - \omega_0}
    \end{align*}

\end{example}
\section{The Wave Equation}
